PoPoolation Pipeline

### 2021-11-08 RADseq.Rmd
Cleaned up RADseq

* vcftools(link)
* gnuplot

@@

## Considerations During Lab Work -- Setting up for Success {#labwork}

@@

## Principles for Analyzing Your Data

### Steps for a robust RAD analysis 
This is just one approach for working through your data. Some people will prefer to run just a subset of samples through a pipeline at first and evaluate parameters, then run all the samples through. Either way, be prepared to make MULTIPLE assemblies and go through this process iteratively, especially if this is a new-to-you study system.

@@

### First, look at the raw data! {#fastqc}  
Always look at your data with fastqc(link) before starting an assembly. First, this is a good check to just make sure the sequencing worked. If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.  
Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
Do you have a lot of adapter sequences? Are the ends of your reads very low quality? If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly. If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.

@@

### Run an assembly pipeline {#assembly}

@@

#### iPyrad walkthrough

Link to annotated ipyrad notebook

### Evaluate potential confounding factors {#confound}

@@

#### "Bad" samples {#bad}

@@

**Missingness in R with SeqArray**
```{r}
#using previously loaded gdsin object
print("Per variant: ")
summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
print("Per sample: ")
summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
samples <- SeqArray::seqGetData(gdsin, "sample.id")
cbind(samples,m2)[order(-m2),]
#plot histogram 
hist(m2,breaks=50)

# close the GDS file
seqClose(gdsin)
```

Another common method of filtering and evaluating you data without ever using R is VCFtools(link). 
**Missingness with vcftools, on the command line**  
```{bash, eval = T}
vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
# sort the file by most missing data and print the top 10 samples  
cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
```
You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.

@@

#### The power of PCA

@@

```{r}
#open the gds file previously created
filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
gdsin = SeqArray::seqOpen(filename.gds)
# exclude samples previously identified as having too much missin g data
bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

@@

Now we will actually run the PCA, again removing the samples with missing data, keeping unlinked SNPs, and removing SNPs with less than 5% minor allele frequency. 
```{r}
# PCA only on SNPs with a minor allele freq greater than 2.5%
pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep) # filtering for pruned SNPs
#close the gds file (saves memory)
#seqClose(gdsin)

@@

Data Info: Below we use RAD data from an Australian marine species complex.
```{r, eval =F}

@@

```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
      ggplot(., aes(x=V1, y=V2, label = ID)) +
      geom_text(size =1.5) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA with Labels")
)
```

```{r}
# which samples have PC1 > -0.7 and PC2 < -0.3?
as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
  filter(V1 > 0.7 & V2 < -0.3) %>%
  select(ID,V1,V2)
```

* Is there no clustering at all? If so, it may be due to:  
  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
  + issues with how your PCA is treating missing data
  + or you actually have no structure in your dataset!
  + `r colorize("figure example, no clustering", "red")`

@@

Here is an example of a PCA with sample replicates, and then with sample replicates removed:

When removing a sample replicate, you can choose one at random or pick the replicate with the least missing data(link). 
```{r, eval=FALSE}
# example code for removing samples from a SNPRelate PCA 
# assumes you have already loaded a vcf file as a GDS
samples.to.remove = c("SampleA_rep","SampleB_rep")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

# PCA only on SNPs with a minor allele freq greater than 2.5%, only keeping samples in keep
pca.out.noreps= SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=1, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep)
#close gds file (saves memory)

@@

#### Visually inspect alignments {#align}  
Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses. In ipyrad this can be done by searching the .loci file, in dDocent... If the alignment looks wonky, you can feel better about removing the whole locus (?).  
`r colorize("code example .loci", "red")`
`r colorize("code dDocent", "red")`  
`r colorize("code stacks", "red")`

#### Batch effects {#batch}
As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage. Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp. when using pre-existing data. Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.

@@

Let's pretend like we have batch effects that were observed on PC 1. With SNPRelate we would extract the PC loadings for each SNP at each PC, then 
```{r}

SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
#load.df <- cbind(SnpLoad$snp.id,SnpLoad$snploading)
#colnames(load.df) <- c("Snp.ID",)

@@

Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them. If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.

#### Cryptic species/contamination/clones {#cryptic}  
One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location. If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses. A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).   
`r colorize("PCA figure of cryptic species", "red")`

Note: Notice the huge variation explained along the first PC-axis.

You can also check heterozygosity, and missingness!

```{r}
```

 Note: Two outliers are obvious. Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples. These samples were later determined using mtDNA and morphological analysis to be cryptic species.
*Unpublished data from a marine species in Australian waters.*

@@

Cryptic clones can also dramatically throw off population genetic analyses. In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates. Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. *other ways to tell they are clones*  
`r colorize("PCA figure of cryptic clones", "red")`

### Test a range of key parameters {#test}

#### Clustering threshold (de novo assembly) {#cluster}  
During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus. This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks). To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs. All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together. If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.  
Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.

@@

#### Mapping parameters (reference-based and some de novo assemblers)   
When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs. In many cases, these SNPs can be filtered out later (link to filtering). However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.

@@

#### Read depth threshold {#seq}

@@

### Filter your data many ways (and embrace the missingness!)
* Don't over filter for missingness
 + papers showing how too much filtering can skew results  
  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021

@@

## Special Cases  
### DARTseq  
### 2bRAD

### Reference-based vs. de novo RADseq
<!---brief outline on general pipeline--->

## Good tutorials  
* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)

>>>>

* [vcftools](https://vcftools.github.io/man_latest.html)
* gnuplot

@@

# Considerations During Lab Work {#labwork}

@@

# Principles for Analyzing Your Data

## Steps for a robust RAD analysis 
This is just one approach for working through your data. Some people will prefer to run just a subset of samples through a pipeline at first and evaluate parameters, then run all the samples through. Either way, be prepared to make MULTIPLE assemblies and go through this process iteratively, especially if this is a new-to-you study system.

@@

## First, look at the raw data! {#fastqc}  
Always look at your data with fastqc(link) before starting an assembly. First, this is a good check to just make sure the sequencing worked. If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.  
Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
Do you have a lot of adapter sequences? Are the ends of your reads very low quality? If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly. If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.

@@

## Run an assembly pipeline {#assembly}

@@

## Evaluate potential confounding factors {#confound}

@@

### "Bad" samples {#bad}

@@

**Missingness in R with SeqArray**
```{r}
#using previously loaded gdsin object
print("Per variant: ")
summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
print("Per sample: ")
summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
samples <- SeqArray::seqGetData(gdsin, "sample.id")
cbind(samples,m2)[order(-m2),]
#plot histogram 
hist(m2,breaks=50)
```

Another common method of filtering and evaluating you data without ever using R is [VCFtools](https://vcftools.github.io/man_latest.html). 
**Missingness with vcftools, on the command line**  
```{bash, eval = T}
vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
# sort the file by most missing data and print the top 10 samples  
cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
```
You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.

@@

### The power of PCA

@@

```{r}
#open the gds file previously created, if not already open
#filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
#gdsin = SeqArray::seqOpen(filename.gds)
# exclude samples previously identified as having too much missin g data
bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

@@

Now we will actually run the PCA, again removing the samples with missing data, keeping unlinked SNPs, and removing SNPs with less than 5% minor allele frequency. 
```{r}
# PCA only on SNPs with a minor allele freq greater than 5%
pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep) # filtering for pruned SNPs
#close the gds file (saves memory)
#seqClose(gdsin)

@@

<!-- Data Info: Below we use RAD data from an Australian marine species complex. ---> 
```{r, eval =F,include=F}

@@

```{r, eval=T}
#PLOT PCA
#PC1 v PC2 with Sample Labels
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
      ggplot(., aes(x=V1, y=V2, label = ID)) +
      geom_text(size =3) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA with Labels")
)
```

```{r, eval=F}
# which samples have PC1 > -0.7 and PC2 < -0.3?
as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
  filter(V1 > 0.7 & V2 < -0.3) %>%
  select(ID,V1,V2)
```

* Is there no clustering at all? If so, it may be due to:  
  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
  + issues with how your PCA is treating missing data
  + or you actually have no structure in your dataset!

<!--`r colorize("figure example, no clustering", "red")`-->

@@

<!--Here is an example of a PCA with sample replicates, and then with sample replicates removed: -->

When removing a sample replicate, you can choose one at random or pick the replicate with the least [missing data](#bad). 
```{r, eval=FALSE}
# example code for removing samples from a SNPRelate PCA 
# assumes you have already loaded a vcf file as a GDS
samples.to.remove = c("SampleA_rep","SampleB_rep")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

# PCA only on SNPs with a minor allele freq greater than 2.5%, only keeping samples in keep
pca.out.noreps= SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=1, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep)

@@

<!-- #### Visually inspect alignments {#align}  
Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses. In ipyrad this can be done by searching the .loci file, in dDocent... If the alignment looks wonky, you can feel better about removing the whole locus (?).  
`r colorize("code example .loci", "red")`
`r colorize("code dDocent", "red")`  
`r colorize("code stacks", "red")`    
-->

### Batch effects {#batch}
As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage. Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp. when using pre-existing data. Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.

@@

In SeqArray/SNPRelate, if you have batch effects that are identifiable by PCA, you would extract the PC loadings for each SNP atthe PC which represents the batch effect, then filter those SNPs out and create a new VCF for downstream analyses. 
```{r, eval=T}
# assume you have a GDS object called gdsin and a PCA object called pca.out
# extract PCA loadings
SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
names(SnpLoad)
dim(SnpLoad$snploading)
#plot a histogram of absolute PC loading values for PC1, with larger loading meaning more effect on PC1
pc<-1
hist(sort(abs(SnpLoad$snploading[pc,]),decreasing=T,index.return=T)[[1]],breaks = 30,main="PC loadings: PC1",)
```
Let's remove the largest effect loci over abs(0.010) and see if it changes our PCA
```{r}
#extract SNP.IDs for loadings on PC 1 > abs(0.015) 
batch.snp.ids <- SnpLoad$snp.id[which(abs(SnpLoad$snploading[1,]) >= 0.010)]

#rerun PCA
snp.id.allfilt <- setdiff(snpset.id,batch.snp.ids)
pca.out.pcfilt = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
                               snp.id=snp.id.allfilt,
                               sample.id = keep) # filtering for pruned SNPs
print(
  as.data.frame(pca.out.pcfilt$eigenvect) %>%
      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
      geom_point(size=2) +
      stat_ellipse(level = 0.95, size = 1) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA Colored by Collection Location")
)

```
To save your filtered GDS object as a VCF: 
```{r}
# set a filter to exclude samples with missing data, and SNPs in linkage or with a batch effect
seqSetFilter(gdsin, sample.id=keep,variant.id=snp.id.allfilt)

# convert to vcf
seqGDS2VCF(gdsin, "RAD_data/OL-miss-pc1-linkage-filt.vcf.gz")

#close the gds file to save memory
closefn.gds(gdsin)

@@

### Cryptic species/contamination/clones {#cryptic}  
One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location. If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses. A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).

<!--Note: Notice the huge variation explained along the first PC-axis.-->

You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.

<!--Note: Two outliers are obvious.
Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples. These samples were later determined using mtDNA and morphological analysis to be cryptic species.
*Unpublished data from a marine species in Australian waters.*-->

@@

Cryptic clones can also dramatically throw off population genetic analyses. In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates. A calculation of sample pairwise distance (e.g., Manhattan distance) can very duplicates/clones.

<!--Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. 
`r colorize("PCA figure of cryptic clones", "red")`-->

## Test a range of key parameters {#test}

### Clustering threshold (de novo assembly) {#cluster}  
During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus. This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks). To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs. All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together. If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.  
Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.

@@

### Mapping parameters (reference-based and some de novo assemblers)   
When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs. In many cases, these SNPs can be filtered out later (link to filtering). However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.

@@

### Read depth threshold {#seq}

@@

## Filter your data many ways
### (and embrace the missingness!) {.unlisted .unnumbered}
* Don't over filter for missingness
 + papers showing how too much filtering can skew results  
  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021

@@

<!--## Special Cases  
### DARTseq  
### 2bRAD

### Reference-based vs. de novo RADseq
-->

# Helpful tutorials  
* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)
